---
title: 'Analyzing COVID-19 Data'
author: "Justin Mende"
date: "6/2/2025"
output: html_document
---

# Using R to Analyze COVID-19 Data
This .Rmd file, as well as the R file found in the same folder,
contain the same overall R code. However, this file was created
to gain experience using and written .Rmd files. It takes in a 
csv file of COVID-19 Data, such as positive cases, tests, and deaths.

Topics used in this file include:
- Reading csv Files 
- Data Frames
- Using the dplyr package
- Filtering and summarizing Data
- Creating plots using plotly and ggplot2
- Creating matrices

The csv file used in this file was retrieved from the following link:
<https://www.kaggle.com/datasets/lin0li/covid19testing?resource=download>

First, the necessary libraries were listed.
```{r}
# Libraries and packages that will be used
library(dplyr)
library(tibble)

library(ggplot2)
library(plotly)
```

Next, the csv file was donwload and stored. It was read in as a data frame.
```{r}
# Get csv data
csv_name <- "analyzing_covid_trends\\tested_worldwide.csv"

covid_data <- read.csv(csv_name)

# Display the head and tail of the data
head(covid_data)
```

A summary table of the data was also created.
```{r}
# Show summary table of data
summary(covid_data)
```

Another number of different functions were used to gain exposure into how they
work and to get key dimensions and features of the data frame.
```{r}
# List the column names
colnames(covid_data)

# Use glimpse to display data
glimpse(covid_data)

# Show list of unique countries listed in the Data
unique(covid_data$Country_Region)

# List the countries in alphabetically order
sort(unique(covid_data$Country_Region))

# Get dimensions of the data frame
dim(covid_data)
nrow(covid_data)
ncol(covid_data)
str(covid_data)
```

The data was then filtered to only get rows involving the entire country. 
Additionally, the names of hte column were changed, and the NA values were
replaced with 0.
```{r}
# Filter data to only get rows where it involves the entire country
country_covid_data <- covid_data %>% filter(Province_State == "All States")

# Remove province column
country_covid_data <- country_covid_data %>% select(-Province_State)
head(country_covid_data)

# Change column names
colnames(country_covid_data) <- c("Date", "Country", "Positve", "Active",
                                  "Hospitalized", "Hospitalized_Currently",
                                  "Recovered", "Death", "Total_Tested",
                                  "Daily_Tested", "Daily_Positive")


# Replace NA values with 0
country_covid_data <- country_covid_data %>% replace(is.na(.), 0)
head(country_covid_data)
```

The data for only Canada was retrieved from the original data frame to be used 
to create different types of graphs.
```{r}
# Filter data to only get the data for Canada
canada_data <- country_covid_data %>% filter(Country == "Canada")
canada_data <- canada_data %>% select(-Country)
canada_data$Date <- as.Date(canada_data$Date)

glimpse(canada_data)
```

First, the plot() function was used to create a line graph of daily positive cases
by date.
```{r}
# Create line graph of daily cases by date
plot(canada_data$Date, canada_data$Daily_Positive, type = "l", lwd = 2,
     main = "Graph of Canada Daily Positive COVID Cases", xlab = "Date",
     ylab = "Number of Cases")
```

Next, using the filtered Canada Data, a ggplot2 line graph was created to display 
active cases by day.
```{r}
# Use ggplot2 to create line graph of active cases by day
ggplot(canada_data, aes(x = Date, y = Active)) + geom_point(color = "blue") +
  labs(title = "Canada: Active COVID-19 Cases")
```

Lastly, a plotly bar graph was created by show deaths by day.
```{r}
# Create bar graph of deaths by day using plotly
canada_bar_graph <- plot_ly(canada_data, x = ~Date, y = ~Death, type = "bar",
                            marker = list(color = "red"))

canada_bar_graph <- canada_bar_graph %>% layout(title = "Canada COVID Deaths",
                                                bargrap = 0.2)

canada_bar_graph
```

The data was furthered summarized by total cases, tests, and hospitalizations by
country.
```{r}
# Summarize data by total tests, cases, hospitalizations by country
country_daily_summary <- country_covid_data %>%
  group_by(Country) %>%
  summarise(Tested = sum(Daily_Tested), Positive = sum(Daily_Positive),
            Hospitalized = sum(Hospitalized_Currently))

country_daily_summary <- country_daily_summary %>% arrange((desc(Positive)))
country_daily_summary
```

Afterwards, to complete the analysis, the top three countries by total positive cases,
tests, hospitalizations and positive test rate were determined and stored in a matrix.
```{r}
# Get top three countries by positive cases
most_positive <- country_daily_summary
most_positive <- head(most_positive, 3)
top_positive_countries <- most_positive$Country
top_positive_countries

# Get top three countries by total tests
most_tested <- country_daily_summary %>% arrange((desc(Tested)))
most_tested <- head(most_tested, 3)
top_tested_countries <- most_tested$Country
top_tested_countries

# Get top three countries by hospitalizations
most_hospitalized <- country_daily_summary %>% arrange((desc(Hospitalized)))
most_hospitalized <- head(most_hospitalized, 3)
top_hospitalized_countries <- most_hospitalized$Country
top_hospitalized_countries

# Calculate testing rate
country_daily_summary$Rate <- country_daily_summary$Positive /
  country_daily_summary$Tested
head(country_daily_summary)

# Get top three countries with highest positive test rate
high_rate <- country_daily_summary %>% arrange((desc(Rate)))
high_rate <- head(high_rate, 3)
highest_rate_countries <- high_rate$Country
highest_rate_countries

# Create matrix with the leaders in each category
covid_leaders <- rbind(top_positive_countries, top_tested_countries,
                       top_hospitalized_countries, highest_rate_countries)
covid_leaders
```